Rapid technological progress in white light-emitting diodes and its sources in innovation and technology spillovers

### 2024-03-30 2_SSL_NE/article/article_main.tex
more edits

% number all lines as pre journal submission guidelines
\usepackage{lineno}
%\linenumbers

@@

Improvements in technology cost and performance, which in this study we generally call “innovations”, can occur via different mechanisms, including (but not limited to) learning by doing\cite{WRIGHT_1936, Arrow_1962}, targeted research and development (R\&D) efforts, or economies of scale\cite{johansson2012global, iea2020perspectives}. The role of these mechanisms at different stages of technology life cycle, from research and development to demonstration, market formation and diffusion\cite{grubler2012policies}, is an area of active research \cite{kavlak2018evaluating,Ziegler2021}. Different mechanisms can result in different kinds of innovations, characterized by what changes in the technology (e.g., its architecture, materials, components, or manufacturing process) and the scale of resulting improvements (e.g., radical or incremental) \cite{Acemoglu_2022}. For example, R\&D efforts (sometimes also called “learning-by-researching”) \cite{Castelnuovo_2005} more often than other mechanisms can result in substantial or even radical improvements in technology performance. On the other hand, learning-by-doing is typically associated with continuous incremental improvements in the technology resulting from accumulated experience in technology demonstration, manufacturing, and use. Both mechanisms, as well as economies of scale, can also lead to radical (i.e., discontinuous) or incremental reductions in technology costs.

Importantly, different mechanisms of innovation in a particular technology play out within a broader innovation system comprising multiple actors, networks, and institutions\cite{grubler2012policies, Anadon2016}. As a result, these mechanisms are affected by various drivers and barriers, including forces of supply and demand\cite{Mowery1979}, public policies, and regulation\cite{anadon2009policy}.

Among these drivers, the role of external knowledge and technology spillovers in research and development of energy technologies remains insufficiently studied \cite{Stephan2021}. While the exact definition of spillovers in the literature depends on the context \cite{Nemet2012, kolesnikov2021spillovers}, we specifically follow the approach of two previous studies \cite{Stephan2021, kolesnikov2021spillovers}. We define technology spillovers as the application of external knowledge in a technology where knowledge is considered external if it has been developed for application in other technologies, sectors, or scientific disciplines. There is emerging evidence that understanding spillovers and the knowledge network beyond a particular technology may be an important factor in understanding \cite{Pichler2020, iea2020cleanenergy} and shaping \cite{Clark2016, Stephan2021} the future evolution of technologies. For example, spillovers can provide critical external knowledge inputs to the R\&D process \cite{kolesnikov2021spillovers}, enable the re-use of experience from different sectors in learning-by-doing\cite{iea2020cleanenergy}, and lead to cost reductions, e.g., via repurposing the manufacturing equipment from other sectors\cite{Stephan2021}.

The aim of this paper is to identify and, where possible, quantify the contribution of different mechanisms and technology spillovers to the historical progress in a specific technology – white light-emitting diode (LED) lighting. Among demand-side technologies, general illumination is a particularly important area for climate change mitigation efforts, as it currently accounts for 15-19\% of global electricity consumption \cite{Zissis2016,doe_electricity}. It is also an area of rapid recent technological change: since the introduction of first commercial white LEDs in 1996 \cite{Nakamura2013}, lighting technology has experienced dramatic efficiency improvements. As shown in \cref{fgr:history_efficacy}, thanks to the introduction of LED-based solid-state lighting (SSL), the efficacy of lighting sources has increased by three orders of magnitude in just over 20 years, which is significantly faster than the historical progress observed in previous lighting technologies \cite{weinold2021quantifying}. For comparison, the highest performing light-emitting devices today reach efficacies of 220 lm/W \cite{lumistrips2021mid}, while an incandescent light bulb can only reach efficacies of up to 18 lm/W. Moreover, this rapid improvement in efficiency has been accompanied by a similarly impressive decrease in LED manufacturing costs and retail prices. \cref{fgr:cost_lamp_small} shows how LED retail prices have fallen by two orders of magnitude, at an annual price per flux decline of 27.3\% during the 2008-2020 period, in line with previous estimates \cite{Gerke2020}.

\begin{figure}[h!]
 \centering
 \includegraphics[width=\textwidth]{figures/history_efficacy.pdf}
 \caption{\textbf{Historical progress in the luminous efficacy ($\eta$) of the most widely-used lighting technologies in human history.} Data points indicate best performers by year of market introduction. Luminous efficacy is the measure of how efficiently a light source converts electrical energy into visible light that can be perceived by the human eye, taking into account the wavelength sensitivity of the eye (see Supplementary Note 2 for details). Dashed lines are guides to the eye, based on a 3rd-order polynomial fit to the data trend. The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_{max}^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al. \cite{Murphy2012}. The magnified plot shows the progress in cool white LEDs from 1996 to 2020, with the dashed line indicating a linear rate of efficacy improvement of 10lm/W per year. For comparison, efficacies of best performers in legacy lighting technologies for 2020 are shown as coloured horizontal lines. Note the logarithmic scale of the vertical axis on the main plot and the linear scale on the magnified plot. Abbreviations: HID - High-Intensity Discharge; CFL - Compact Fluorescent Lamp; Hal. - Halogen, Incd. - Incandescent. Source: own synthesis of published data based on a visual approach proposed by Azevedo et al. \cite{azevedo2009transition}. See Supplementary Note 16 for the full list of sources and references.}
 \label{fgr:history_efficacy}
\end{figure}

\begin{figure}[h!]
\centering
  \includegraphics[height=5cm]{figures/cost_lamp_small.pdf}
  \caption{\textbf{Historical development of retail sales prices per luminous flux for LED-based luminaires in the period 2008-2020.} Luminaires considered include light bulbs, spotlights and recessed lights. Red curved and dashed lines represent average retail sales prices and price projections for LED based luminaires published by the U.S. Department of Energy (DOE) \cite{national2013assessment}. Shown for reference are the average prices for compact fluorescent (CFL) and incandescent light bulbs, with the latter assumed constant based on the average in the covered time period. Note that projections are consistently at the lower end of the retail price data. This is because retail price data includes more expensive luminaire types, while projections are for traditional light bulbs only. Projections also do not include value added tax. Source: own synthesis of data on sales prices collected from various consumer watchdog databases and publications. See Supplementary Note 16 for the full list of sources and references.}
  \label{fgr:cost_lamp_small}
\end{figure}

These dramatic improvements in lighting technology, supported by the introduction of lighting efficiency regulations phasing out incandescent light bulbs and targeted policies stimulating LED adoption in many countries, led to the rapid expansion and diffusion of SSL technologies \cite{weinold2020long, stegmaier2021incandescent, Mills2014}. As a result, by 2020, highly efficient LED luminaires were saving an estimated 131 TWh/year in the EU \cite{eu2019impactass} and 442 TWh/year in the US \cite{guidehouse2020adoption}, which is on par with the amount of energy produced in the same year by all photovoltaic installations in these regions. Notably, market adoption of LED lighting is not limited to developed economies \cite{Kamat2020}. For example, durability, low up-front cost and high efficiency of LED light sources have led to early and widespread adoption in rural West African communities without access to grid electricity \cite{Bensch2017}. LEDs have also been used in a wide range of applications beyond lighting, such as personal health monitors \cite{Wyatt2020}, potable water treatment \cite{Lui2014}, high-bandwidth wireless data transmission \cite{Haas2016}, and augmented reality eye wear \cite{Lee2016}.

@@

To address these questions, we first identify three sets of metrics to quantify historical progress in LED lighting technology: 1) metrics of energy efficiency of LED devices, including the total device efficiency (“lamp efficiency”) and the sub-efficiencies that describe different energy loss channels in a LED device; 2) metrics of lighting quality relevant to consumer experience, specifically the colour rendering index (CRI) and colour temperature (CT); and 3) LED chip manufacturing cost. The rationale for the choice of these metrics, along with their detailed descriptions and definitions, are discussed in Supplementary Notes section 2. Next, using data from literature review, patent analysis, and expert interviews, we trace the historical improvements in these metrics in warm white GaN-based phosphor-converted LEDs—currently the most widely adopted variant of white LED technology which we thus choose as the focal technological area for our analysis—from their introduction to the lighting market in 2003 to 2020, the year with the most recent data available at the time of data collection. We find that total LED device efficiency of such LEDs increased from 5.8\% to 38.8\% between 2003 and 2020. A majority of this increase can be attributed to a series of innovations in white LED technology resulting from R\&D efforts, with at least 8.5\% of the cumulative increase attributed specifically to technological innovations driven by technology spillovers. Our interviews also revealed a lesser but still important role of incremental manufacturing process improvements most likely resulting from learning-by-doing. However, substantial further research is needed to quantify the relative contribution of learning-by-doing vis-a-vis technological innovations to this progress. On the other hand, we find that innovations driven by technology spillovers contributed to nearly all consumer experience improvements in white LED lighting technology, thus playing an important role in widespread adoption of SSL in general illumination applications. Finally, our bottom-up LED manufacturing cost model shows a 95.5\% decrease in the cost of producing low-to-mid-power classic-chip-architecture GaN-based phosphor-converted warm-white LEDs from 1.11\$ to 0.05\$ (in 2020 USD) between 2003 and 2020. In contrast with LED performance improvements, where progress has been driven mostly by R\&D efforts, the dramatic decline in LED manufacturing cost has been a product of learning-by-doing resulting in higher yields across manufacturing steps and economies of scale resulting from increases in the sapphire wafer size. We provide the details of our findings for the progress in white LED efficiency, cost, and consumer experience, as well as the role of technology spillovers in this progress, in the remaining sections below.

@@

Using the mixed-methods approach described in Methods, we have collected data on the historical progress in individual sub-efficiencies of warm white phosphor-converted GaN-based LED devices, along with white LED innovations that have driven this progress. The results of our data collection efforts are presented in Supplementary Figure 7 and discussed in Supplementary Notes section 4. Using this information, we calculate the overall lamp efficiency $\eta_L$ for the best performing mid/high-power phosphor-converted warm white LED devices in four years: 2003, 2010, 2016 and 2020. The waterfall diagrams of electric power input losses in \cref{fgr:waterfall} show how improvements in individual sub-efficiencies led to improvements in the overall white LED lamp efficiency from $\eta_L=5.8\%$ in 2003 to 12.7\% in 2010, 32.5\% in 2016 and finally to 38.8\% in 2020. As is evident from the figure, no single loss channel dominates in terms of its contribution to the overall efficiency, in line with previous observations \cite{tsao2010solid}. We note, however, that the loss channels with a fixed physical limit on efficiency, e.g., Stokes loss that determines the light conversion efficiency by phosphors, became relatively more dominant in 2016 and 2020 compared to 2003 and 2010.  This is a direct result of large efficiency improvements in upstream sub-efficiencies.

\cref{fgr:breakthroughs_efficiency} shows the overall magnitude of contributions of identified LED innovations and technology spillovers to improvements in LED efficiency over time across different sub-efficiencies. The full list of identified LED innovations considered in our study is provided in Table SI2 in Section 3 of the Supplementary Information document. The list of corresponding technology spillovers is provided in \cref{tab:spillovers}. Through the index decomposition analysis described in Section 2.5 in the Supplementary Information we find that out of the overall LED efficiency increase of 32.9\% from 5.8\% to 38.8\% between 2003 and 2020, at least 2.8 percentage points can be attributed specifically to innovations driven by technology spillovers identified in this study, corresponding to 8.5\% of the total LED efficiency improvements between 2003 and 2020.

In \cref{fgr:breakthroughs_efficiency} we also compare, for the first time, efficiency improvements across sub-efficiencies over time, contrasting them with the physical limits of the corresponding loss channels. The estimates shown here for the contributions of different channels do not show uncertainty ranges because we lack reliable information about the distribution of input parameters, which would require us to arbitrarily assign range values for inputs, and the fact that we focus on a best case scenario. Thus, our point estimates should be interpreted as the expected values in a best case scenario. We find that there has been consistent progress across all device sub-efficiencies in the recorded period. Specifically, between 2003 and 2020, forward voltage efficiency increased from 70\% to 99.5\%, internal quantum efficiency from 55\% to 90\%, electrical droop from 65\% to 90\%, light extraction efficiency from 60\% to 90\%, spectral efficiency from 74\% to 83\%, conversion efficiency (red) from 11\% to 45\%, conversion efficiency (green) from 19\% to 61\%. Notably, some sub-efficiencies for the most recent devices considered in our study are now within $\sim10\%$ of their respective physical limits. The exception is spectral efficiency which, at $\sim17\%$ below the physical limit, shows larger potential for further improvements, which is important, given that efforts to improve white LED performance across different efficiency loss channels through R\&D continue \cite{cho2017white, Weisbuch2020}.

\begin{figure}[H]
 \centering
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2003.pdf}
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2010.pdf}
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2016.pdf}
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2020.pdf}
 \captionsetup{font=footnotesize}
 \caption{\textbf{Waterfall diagrams of the loss channels in a generic mid/high-power white LED package for 2003, 2010, 2016 and 2020}. Losses are normalized to 1W of electric power input. The sub-efficiencies corresponding to each loss channel are listed below each column and described in Supplementary Note 2. Numbers for each loss channel indicate energy losses both in relative terms of input power (in percent) at the point of the channel and absolute values (in Watts). Percentages for loss channels labeled by red, green and blue font indicate losses of corresponding remaining red/green/blue light energy. The following LED architectures and light-source spectra used in calculations for each considered year are shown for reference: 2003 – Flip-chip with YGAG phosphor; 2010 – Flip-chip with 258 phosphor; 2016 – Flip-chip with PFS phosphor; 2020 – Chip-scale package flip-chip with SALON phosphor. Note that the inset spectra clearly show progress in spectral efficiency. The red curve shows the wavelength-dependent sensitivity of the human eye, while the black curve shows the spectrum of the light source. When the spectrum has significant energy at the edges of the sensitivity curve, energy is lost. Details for each architecture are provided in \cref{fgr:chip_architecture_overview}.  Details on the phosphors for light down conversion are provided in \cref{tab:phosphors} and Supplementary Note 4. Abbreviations: Scat. = Scattering; Nonrad. = Nonradiative. Source: own elaboration based on data on sub-efficiencies presented in Supplementary Figure 9 and spectral data in Supplementary Figure 10, adapted from publications on the respective phosphors: YGAG (2003)\cite{Mueller2002}, 258 (2010)\cite{MuellerMach2005}, PFS (2016)\cite{Murphy2015}, SALON (2019)\cite{Hoerder2019}.}
 \label{fgr:waterfall}
\end{figure}

\begin{figure}[H]
 \centering
 \includegraphics[width=\textwidth]{figures/breakthroughs_efficiency.pdf}
 \caption{\textbf{Contribution of innovations and technology spillovers to the historical progress in sub-efficiencies of phosphor-converted warm white LEDs.} LEDs with test currents of at least 350mA are considered. The following sub-efficiencies are represented: $\eta_{V_f}$ - forward voltage efficiency; $\eta_IQ$ - internal quantum efficiency; $\eta_{droop}$ - efficiency droop; $\eta_{LE}$ - light extraction efficiency; $\eta_{C_R}$ - conversion efficiency for red phosphors; $\eta_{C_{Y/G}}$ - conversion efficiency for yellow/green phosphors; $\eta_S$ - spectral efficiency. The overall LED lamp efficiency $\eta_L$ is displayed in the rightmost column. Vertical bars represent cumulative contributions to efficiency improvements from LED technology innovations, with purple bars indicating innovations driven by technology spillovers (annotated and listed in an inset table) and grey bars indicating all other improvements identified in this study. Horizontal coloured lines indicate sub-efficiency levels of best performing LEDs for the four years used in \cref{fgr:waterfall}: 2003, 2010, 2016, and 2020. For additional historical context, data for 1997 is included whenever possible. "N/A" denotes sub-efficiencies where 1997 data could not be calculated for the following reasons: $\eta_{V_f}, \eta_{droop}$ depend on the device current, which was below 350mA in 1997, making a comparison with contemporary devices difficult. $\eta_{C_R}, \eta_{C_{Y/G}}$ and $\eta_S$ are relevant only to warm white spectrum LEDs, which were not available in 1997. Note: ITO current spreading layer affects different sub-efficiencies in different chip architectures, e.g., in modern flip-chip architectures $\eta_{LE}$ no longer depends on ITO, see \cref{fgr:chip_architecture_overview}. Physical limits on sub-efficiencies are indicated by black horizontal lines. Note that the physical limit of $\eta_{V_f}$ is above 100\% due to quantum effects \cite{david2016electrical}. Source: own elaboration based on data represented in \cref{fgr:waterfall}, \cref{tab:spillovers} and Supplementary Note 4.}
 \label{fgr:breakthroughs_efficiency}
\end{figure}

@@

Historical improvements in consumer experience metrics for phosphor-converted warm white GaN-based LEDs are shown in \cref{fgr:consumer_experience}. In general illumination applications, a high colour rendering index (CRI) in combination with a specific, tunable range of possible colour temperatures is desirable. Both metrics are determined by LED device emission spectra, which, in turn, depend on the properties of materials used for the conversion of blue light generated by conventional GaN-based LEDs into the white light required for general illumination . It allows us to establish the links between all major improvements in the two consumer experience metrics considered in this study and individual LED innovations  associated either with phosphors or quantum dots used for light down conversion in LEDs. The first commercial white LED produced by Nichia in 1996 used a YAG (Yttrium Aluminium Garnet) phosphor that could generate only cool white light \cite{bando1998development}. As shown in \cref{fgr:consumer_experience}, after a series of innovations listed in \cref{tab:phosphors}, LEDs today can be tuned for high CRI performance and a range of desirable colour temperatures. All innovations in this list are primarily a result of R\&D efforts.

>>>>

% number all lines as pre journal submission guidelines
\usepackage{lineno}
\linenumbers

@@

Improvements in technology cost and performance, which in this study we generally call “innovations”, can occur via different mechanisms, including (but not limited to) learning by doing \cite{WRIGHT_1936, Arrow_1962}, targeted research and development (R\&D) efforts, or economies of scale \cite{johansson2012global, iea2020perspectives}. The role of these mechanisms at different stages of the technology life cycle, from research and development to demonstration, market formation and diffusion \cite{grubler2012policies}, is an area of active research \cite{kavlak2018evaluating,Ziegler2021}. Different mechanisms can result in different kinds of innovations, characterized by what changes in the technology (e.g., its architecture, materials, components, or manufacturing process) and the scale of resulting improvements (e.g., radical or incremental) \cite{Acemoglu_2022}. For example, R\&D efforts (sometimes also called “learning-by-researching”) \cite{Castelnuovo_2005} more often than other mechanisms can result in substantial or even radical improvements in technology performance. On the other hand, learning-by-doing is typically associated with continuous incremental improvements in the technology resulting from accumulated experience in technology demonstration, manufacturing, and use. Both mechanisms, as well as economies of scale, can also lead to radical (i.e., discontinuous) or incremental reductions in technology costs.

Importantly, different mechanisms of innovation in a particular technology play out within a broader innovation system comprising multiple actors, networks, and institutions \cite{grubler2012policies, Anadon2016}. As a result, these mechanisms are affected by various drivers and barriers, including forces of supply and demand \cite{Mowery1979}, public policies, and regulation \cite{anadon2009policy}.

Among these drivers, the role of external knowledge and technology spillovers in research and development of energy technologies remains insufficiently studied \cite{Stephan2021}. While the exact definition of spillovers in the literature depends on the context \cite{Nemet2012, kolesnikov2021spillovers}, we specifically follow the approach of two previous studies \cite{Stephan2021, kolesnikov2021spillovers}. We define technology spillovers as the application of external knowledge in a technology where knowledge is considered external if it has been developed for application in other technologies, sectors, or scientific disciplines. There is emerging evidence that understanding spillovers and the knowledge network beyond a particular technology may be an important factor in understanding \cite{Pichler2020, iea2020cleanenergy} and shaping \cite{Clark2016, Stephan2021} the future evolution of technologies. For example, spillovers can provide critical external knowledge inputs to the R\&D process \cite{kolesnikov2021spillovers}, enable the re-use of experience from different sectors in learning-by-doing \cite{iea2020cleanenergy}, and lead to cost reductions, e.g., via repurposing the manufacturing equipment from other sectors \cite{Stephan2021}.

The aim of this paper is to identify and, where possible, quantify the contribution of different mechanisms and technology spillovers to the historical progress in a specific technology – white light-emitting diodes (LEDs) for general illumination. Among demand-side technologies, lighting is a particularly important area for climate change mitigation efforts, as it currently accounts for 15-19\% of global electricity consumption \cite{Zissis2016,doe_electricity}. It is also an area of rapid recent technological change: since the introduction of first commercial white LEDs in 1996 \cite{Nakamura2013}, lighting technology has experienced dramatic efficiency improvements. As shown in \cref{fgr:history_efficacy}, thanks to the introduction of LED-based solid-state lighting (SSL), the efficacy of lighting sources has increased by three orders of magnitude in just over 20 years, which is significantly faster than the historical progress observed in previous lighting technologies \cite{weinold2021quantifying}. For comparison, the highest performing light-emitting devices today reach efficacies of 220 lm/W \cite{lumistrips2021mid}, while an incandescent light bulb can only reach efficacies of up to 18 lm/W. Moreover, this rapid improvement in efficiency has been accompanied by a similarly impressive decrease in LED manufacturing costs and retail prices. \cref{fgr:cost_lamp_small} shows how LED retail prices have fallen by two orders of magnitude, at an annual price per flux decline of 27.3\% during the 2008-2020 period, in line with previous estimates \cite{Gerke2020}.

\begin{figure}[h!]
 \centering
 \includegraphics[width=\textwidth]{figures/history_efficacy.pdf}
 \caption{\textbf{Historical progress in the luminous efficacy ($\eta$) of the most widely-used lighting technologies in human history.} Data points indicate best performers by year of market introduction. Luminous efficacy is the measure of how efficiently a light source converts electrical energy into visible light that can be perceived by the human eye, taking into account the wavelength sensitivity of the eye (see Supplementary Note 2 for details). Dashed lines are guides to the eye, based on a 3rd-order polynomial fit to the data trend. The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_{max}^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al. \cite{Murphy2012}. The magnified plot shows the progress in cool white LEDs from 1996 to 2020, with the dashed line indicating a linear rate of efficacy improvement of 10lm/W per year. For comparison, efficacies of best performers in legacy lighting technologies for 2020 are shown as coloured horizontal lines. Note the logarithmic scale of the vertical axis on the main plot and the linear scale on the magnified plot. Abbreviations: HID - High-Intensity Discharge; CFL - Compact Fluorescent Lamp; Hal. - Halogen, Incd. - Incandescent. Source: own synthesis of published data based on a visual approach proposed by Azevedo et al. \cite{azevedo2009transition}. See Supplementary Note 8 for the full list of sources and references.}
 \label{fgr:history_efficacy}
\end{figure}

\begin{figure}[h!]
\centering
  \includegraphics[height=5cm]{figures/cost_lamp_small.pdf}
  \caption{\textbf{Historical development of retail sales prices per luminous flux for LED-based luminaires in the period 2008-2020.} Luminaires considered include light bulbs, spotlights and recessed lights. Red curved and dashed lines represent average retail sales prices and price projections for LED based luminaires published by the U.S. Department of Energy (DOE) \cite{national2013assessment}. Shown for reference are the average prices for compact fluorescent (CFL) and incandescent light bulbs, with the latter assumed constant based on the average in the covered time period. Note that projections are consistently at the lower end of the retail price data. This is because retail price data includes more expensive luminaire types, while projections are for traditional light bulbs only. Projections also do not include value added tax. Source: own synthesis of data on sales prices collected from various consumer watchdog databases and publications. See Supplementary Note 8 for the full list of sources and references.}
  \label{fgr:cost_lamp_small}
\end{figure}

These dramatic improvements in lighting technology, supported by the introduction of efficiency regulations phasing out incandescent light bulbs and targeted policies stimulating LED adoption in many countries, led to the rapid expansion and diffusion of SSL technologies \cite{weinold2020long, stegmaier2021incandescent, Mills2014}. As a result, by 2020, highly efficient LED luminaires were saving an estimated 131 TWh/year in the EU \cite{eu2019impactass} and 442 TWh/year in the US \cite{guidehouse2020adoption}, which is on par with the amount of energy produced in the same year by all photovoltaic installations in these regions. Notably, market adoption of LED lighting is not limited to developed economies \cite{Kamat2020}. For example, durability, low up-front cost and high efficiency of LED light sources have led to early and widespread adoption in rural West African communities without access to grid electricity \cite{Bensch2017}. LEDs have also been used in a wide range of applications beyond lighting, such as personal health monitors \cite{Wyatt2020}, potable water treatment \cite{Lui2014}, high-bandwidth wireless data transmission \cite{Haas2016}, and augmented reality eye wear \cite{Lee2016}.

@@

To address these questions, we first identify three sets of metrics to quantify historical progress in LED lighting technology: 1) metrics of energy efficiency of LED devices, including the total device efficiency (“lamp efficiency”) and the sub-efficiencies that describe different energy loss channels in a LED device; 2) metrics of lighting quality relevant to consumer experience, specifically the colour rendering index (CRI) and colour temperature (CT); and 3) LED chip manufacturing cost. The rationale for the choice of these metrics, along with their detailed descriptions and definitions, are discussed in Supplementary Method 2. Next, using data from literature review, patent analysis, and expert interviews, we trace the historical improvements in these metrics in warm white GaN-based phosphor-converted LEDs—currently the most widely adopted variant of white LED technology which we thus choose as the focal technological area for our analysis—from their introduction to the lighting market in 2003 to 2020, the year with the most recent data available at the time of data collection.

We find that total LED device efficiency of such LEDs increased from 5.8\% to 38.8\% between 2003 and 2020. A majority of this increase can be attributed to a series of innovations in white LED technology resulting from R\&D efforts, with at least 8.5\% of the cumulative increase attributed specifically to technological innovations driven by technology spillovers. Our interviews also revealed a lesser but still important role of incremental manufacturing process improvements most likely resulting from learning-by-doing. However, substantial further research is needed to quantify the relative contribution of learning-by-doing vis-a-vis technological innovations to this progress. On the other hand, we find that innovations driven by technology spillovers contributed to nearly all consumer experience improvements in white LED lighting technology, thus playing an important role in widespread adoption of SSL in general illumination applications. Finally, our bottom-up LED manufacturing cost model shows a 95.5\% decrease in the cost of producing low-to-mid-power classic-chip-architecture GaN-based phosphor-converted warm-white LEDs from 1.11\$ to 0.05\$ (in 2020 USD) between 2003 and 2020. In contrast with LED performance improvements, where progress has been driven mostly by R\&D efforts, the dramatic decline in LED manufacturing cost has been a product of learning-by-doing resulting in higher yields across manufacturing steps and economies of scale resulting from increases in the sapphire wafer size. We provide the details of our findings for the progress in white LED efficiency, cost, and consumer experience, as well as the role of technology spillovers in this progress, in the remaining sections below.

@@

Using the mixed-methods approach described in Methods, we collected data on the historical progress in individual sub-efficiencies of warm white phosphor-converted GaN-based LED devices, along with white LED innovations that have driven this progress. The results of our data collection efforts are presented in Supplementary Note 4. Using this information, we calculate the overall lamp efficiency $\eta_L$ for the best performing mid/high-power phosphor-converted warm white LED devices in four years: 2003, 2010, 2016 and 2020. The waterfall diagrams of electric power input losses in \cref{fgr:waterfall} show how improvements in individual sub-efficiencies led to improvements in the overall white LED lamp efficiency from $\eta_L=5.8\%$ in 2003 to 12.7\% in 2010, 32.5\% in 2016 and finally to 38.8\% in 2020. As is evident from the figure, no single loss channel dominates in terms of its contribution to the overall efficiency, in line with previous observations \cite{tsao2010solid}. We note, however, that the loss channels with a fixed physical limit on efficiency, e.g., Stokes loss that determines the light conversion efficiency by phosphors, became relatively more dominant in 2016 and 2020 compared to 2003 and 2010.  This is a direct result of large efficiency improvements in upstream sub-efficiencies.

\cref{fgr:breakthroughs_efficiency} shows the overall magnitude of contributions of identified LED innovations and technology spillovers to improvements in LED efficiency over time across different sub-efficiencies. The full list of identified LED innovations considered in our study is provided in Supplementary Note 5. The list of corresponding technology spillovers is provided in \cref{tab:spillovers}. Through the index decomposition analysis described in Section 2.5 in the Supplementary Information we find that out of the overall LED efficiency increase of 32.9\% from 5.8\% to 38.8\% between 2003 and 2020, at least 2.8 percentage points can be attributed specifically to innovations driven by technology spillovers identified in this study, corresponding to 8.5\% of the total LED efficiency improvements between 2003 and 2020.

In \cref{fgr:breakthroughs_efficiency} we also compare, for the first time, efficiency improvements across sub-efficiencies over time, contrasting them with the physical limits of the corresponding loss channels. Since we focus on a best case scenario, and because we lack reliable information about the distribution of input parameters, which would require us to arbitrarily assign range values for inputs, these estimates do not show uncertainty ranges. Thus, our point estimates should be interpreted as the expected values in a best case scenario. We find that there has been consistent progress across all device sub-efficiencies in the recorded period. Specifically, between 2003 and 2020, forward voltage efficiency increased from 70\% to 99.5\%, internal quantum efficiency from 55\% to 90\%, electrical droop from 65\% to 90\%, light extraction efficiency from 60\% to 90\%, spectral efficiency from 74\% to 83\%, conversion efficiency (red) from 11\% to 45\%, conversion efficiency (green) from 19\% to 61\%. Notably, some sub-efficiencies for the most recent devices considered in our study are now within $\sim10\%$ of their respective physical limits. The exception is spectral efficiency which, at $\sim17\%$ below the physical limit, shows larger potential for further improvements, which is important, given that efforts to improve white LED performance across different efficiency loss channels through R\&D continue \cite{cho2017white, Weisbuch2020}.

\begin{figure}[H]
 \centering
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2003.pdf}
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2010.pdf}
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2016.pdf}
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2020.pdf}
 \captionsetup{font=footnotesize}
 \caption{\textbf{Waterfall diagrams of the loss channels in a generic mid/high-power white LED package for 2003, 2010, 2016 and 2020}. Losses are normalized to 1W of electric power input. The sub-efficiencies corresponding to each loss channel are listed below each column and described in Supplementary Note 2. Numbers for each loss channel indicate energy losses both in relative terms of input power (in percent) at the point of the channel and absolute values (in Watts). Percentages for loss channels labeled by red, green and blue font indicate losses of corresponding remaining red/green/blue light energy. The following LED architectures and light-source spectra used in calculations for each considered year are shown for reference: 2003 – Flip-chip with YGAG phosphor; 2010 – Flip-chip with 258 phosphor; 2016 – Flip-chip with PFS phosphor; 2020 – Chip-scale package flip-chip with SALON phosphor. Note that the inset spectra clearly show progress in spectral efficiency. The red curve shows the wavelength-dependent sensitivity of the human eye, while the black curve shows the spectrum of the light source. When the spectrum has significant energy at the edges of the sensitivity curve, energy is lost. Details for each architecture are provided in Supplementary Note 3.  Details on the phosphors for light down conversion are provided in \cref{tab:phosphors} and Supplementary Note 4. Abbreviations: Scat. = Scattering; Nonrad. = Nonradiative. Source: own elaboration based on data on sub-efficiencies presented in Supplementary Figure 9 and spectral data in Supplementary Figure 10, adapted from publications on the respective phosphors: YGAG (2003)\cite{Mueller2002}, 258 (2010)\cite{MuellerMach2005}, PFS (2016)\cite{Murphy2015}, SALON (2019)\cite{Hoerder2019}.}
 \label{fgr:waterfall}
\end{figure}

\begin{figure}[H]
 \centering
 \includegraphics[width=\textwidth]{figures/breakthroughs_efficiency.pdf}
 \caption{\textbf{Contribution of innovations and technology spillovers to the historical progress in sub-efficiencies of phosphor-converted warm white LEDs.} LEDs with test currents of at least 350mA are considered. The following sub-efficiencies are represented: $\eta_{V_f}$ - forward voltage efficiency; $\eta_IQ$ - internal quantum efficiency; $\eta_{droop}$ - efficiency droop; $\eta_{LE}$ - light extraction efficiency; $\eta_{C_R}$ - conversion efficiency for red phosphors; $\eta_{C_{Y/G}}$ - conversion efficiency for yellow/green phosphors; $\eta_S$ - spectral efficiency. The overall LED lamp efficiency $\eta_L$ is displayed in the rightmost column. Vertical bars represent cumulative contributions to efficiency improvements from LED technology innovations, with purple bars indicating innovations driven by technology spillovers (annotated and listed in an inset table) and grey bars indicating all other improvements identified in this study. Horizontal coloured lines indicate sub-efficiency levels of best performing LEDs for the four years used in \cref{fgr:waterfall}: 2003, 2010, 2016, and 2020. For additional historical context, data for 1997 is included whenever possible. "N/A" denotes sub-efficiencies where 1997 data could not be calculated for the following reasons: $\eta_{V_f}, \eta_{droop}$ depend on the device current, which was below 350mA in 1997, making a comparison with contemporary devices difficult. $\eta_{C_R}, \eta_{C_{Y/G}}$ and $\eta_S$ are relevant only to warm white spectrum LEDs, which were not available in 1997. Note: ITO current spreading layer affects different sub-efficiencies in different chip architectures, e.g., in modern flip-chip architectures $\eta_{LE}$ no longer depends on ITO, see Supplementary Note 3. Physical limits on sub-efficiencies are indicated by black horizontal lines. Note that the physical limit of $\eta_{V_f}$ is above 100\% due to quantum effects \cite{david2016electrical}. Source: own elaboration based on data represented in \cref{fgr:waterfall}, \cref{tab:spillovers} and Supplementary Note 4.}
 \label{fgr:breakthroughs_efficiency}
\end{figure}

@@

Historical improvements in consumer experience metrics for phosphor-converted warm white GaN-based LEDs are shown in \cref{fgr:consumer_experience}. In general illumination applications, a high colour rendering index (CRI) in combination with a specific, tunable range of possible colour temperatures is desirable. Both metrics are determined by LED device emission spectra, which, in turn, depend on the properties of materials used for the conversion of blue light generated by conventional GaN-based LEDs into white light. It allows us to establish the links between all major improvements in the two consumer experience metrics considered in this study and individual LED innovations  associated either with phosphors or quantum dots used for light down conversion in LEDs. The first commercial white LED produced by Nichia in 1996 used a YAG (Yttrium Aluminium Garnet) phosphor that resulted in cool white light only \cite{bando1998development}. As shown in \cref{fgr:consumer_experience}, after a series of innovations listed in \cref{tab:phosphors}, LEDs today can be tuned for high CRI performance and a range of desirable colour temperatures. All innovations in this list are primarily a result of R\&D efforts.

Notably, from detailed descriptions of the history of innovations in this list, provided in Supplementary Note 6, we find that only a single innovation related to LED consumer experience improvements was purposefully developed for application in solid-state lighting: the 2016 SALON phosphor compound \cite{Hoerder2019,seibald2019phosphor}. All other innovations in the list were either originally developed for non-LED applications or prominently used knowledge from areas of science and technology beyond LED or SSL (see also \cref{tab:spillovers}). This means that technology spillovers contributed to nearly all consumer experience improvements in LED-based lighting technology, thus playing an important role in widespread adoption of SSL in general illumination applications.
